MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-04-14 latex/paper/paper.tex
more lutff

    Here we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
    However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
    Configurable logic devices have undergone significant evolution over the past four decades. 
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
    Figure \ref{fig:pla} shows one such PLA architecture.

@@

    \begin{figure}
        \centering
        \includegraphics[width=8.0cm]{figures/homogenous_fpga.png}
        \includegraphics[width=8.0cm]{figures/heterogenous_fpga_3.jpg}
        \caption{
            \textbf{Top}: A homogeneous island-style FPGA architecture with 16 CLBs in a grid. 
            \textbf{Bottom}: A heterogeneous island-style FPGA with a mix of CLBs and macrocells.
        }
        \label{fig:fpga}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
    \end{figure}

@@

    This brings us to modern day FPGA architectures. 
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.

@@

    In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
    The 7-Series architecture first released in 2010 and was superceded in 2013 by the Ultrascale architecture which made several improvements over the 7-Series. 
    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.

    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

@@

    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL. 
    These SLICE Sites facilitate the core programmable logic of the FPGA.

\section{The FPGA Design Flow and Toolchain}

    \begin{figure}
        \centering
        \includegraphics[width=8.0cm]{figures/design_flow.png}
        \caption{A typical FPGA design and verification workflow}
        \label{fig:design_flow}
    \end{figure}

@@

    The process begins with \textbf{design entry}, where an engineer describes the desired functionality using a hardware description language (HDL) such as Verilog or VHDL. 
    Engineers may use various coding styles - behavioral, structural, dataflow, or a mix of all three - to capture the logic and behavior of the design while abstracting away the intricacies of the physical hardware architecture.

@@

        \subsection{LUT-FF Pairs}
            In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}.
            Shown in Figure \ref{fig:intrasite_intersite} are two possible placements for this cell net. 
            On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
            On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.

@@

        \begin{figure}[H]
            \centering
            \includegraphics[width=8.0cm]{figures/intersite_intrasite.png}
            \caption{
                \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
                \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
                \textbf{Bottom}: The LUT-FF pair placed across different Sites.
            }
            \label{fig:intrasite_intersite}
        \end{figure}

@@

>>>>

    First we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
    However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
    Configurable logic devices have undergone significant evolution over the past four decades. 
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
    Figure \ref{fig:pla} shows one such PLA architecture.

@@

    \begin{figure}
        \centering
        \includegraphics[width=8.0cm]{figures/homogenous_fpga.png}
        \includegraphics[width=8.0cm]{figures/heterogenous_fpga_3.jpg}
        \caption{
            \textbf{Top}: A homogeneous island-style FPGA architecture with 16 CLBs in a grid. 
            \textbf{Bottom}: A heterogeneous island-style FPGA with a mix of CLBs and macrocells.
        }
        \label{fig:fpga}
        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
    \end{figure}

@@

    This brings us to modern day FPGA architectures. 
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

@@

    In this paper we will be working with the Xilinx 7-Series architecture which follows a heterogenous island-style design. 
    The 7-Series architecture first released in 2010 and was then superceded in 2013 by the Ultrascale architecture which offered several improvements over the 7-Series. 
    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to hobbyists and compatibility with open source tooling.

    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
    At its most basic level, an FPGA is a vast physical array of atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

@@

    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing a collection of BELs - 8 LUTs, 8 Flip Flops, and 1 CARRY4.

    The LUTs and Flip Flops constitute the core of the FPGA programmable logic.

    Flip Flops (FFs) facilitate synchronous (clock-driven or event-driven) registers.

    SLICE Sites come in two flavors - SLICEL (SLICE-Logic) and SLICEM (SLICE-Memory). 
    In the SLICEL, Lookup Tables (LUTs), as their name implies, are fast shallow read-only memories (ROMs) with 6 input pins and 1 or 2 output pins. 
    All combinational logic - boolean functions - are mapped onto these LUTs as read-only-memories during synthesis. 
    Larger boolean functions with more than 6 inputs will be mapped onto multiple cascaded LUTs.

    The SLICEM is similar to the SLICEL, the key difference being that their LUTs can optionally be programmed as RAMs to facilitate "Distributed RAM", an alternative to the larger dedicated Block RAM (BRAM) macro. 
    Xilinx devices usually come a 25\% SLICEM, 75\% SLICEL split amongst all SLICE Sites, where a CLB Tile can contain either two SLICELs or one SLICEL one SLICEM.

    Each SLICE also comes with a shallow 4-bit carry-lookahead adder (CLA) BEL called "CARRY4".

\section{The FPGA Design Flow and Toolchain}
    \begin{figure}
        \centering
        \includegraphics[width=8.0cm]{figures/design_flow.png}
        \caption{A typical FPGA design and verification workflow}
        \label{fig:design_flow}
    \end{figure}

    In addition to the FPGA hardware architecture, one should also understand the role that placement plays in the borader FPGA design flow and software toolchain.

@@

    The process begins with \textbf{design entry}, where an engineer describes the desired functionality of a digital system using a hardware description language (HDL) such as Verilog or VHDL. 
    Engineers may use various coding styles - behavioral, structural, dataflow, or a mix of all three - to capture the logic and behavior of the design while abstracting away the intricacies of the physical hardware architecture.

@@

        \subsection{LUT-FF Pairs}
            Very often, engineers model the behavior of a digital system as a collection of Finite State Machines (FSM) as shown in Figure \ref{fig:fsm}. 
            An FSM at its core is a synchronous sequential circuit comprising of a combinational element and a synchronous memory element.
            In the 7-Series architecture, combinational elements are synthesized as LUTs, where the truth table of the combinational function is mapped onto the LUT's memory.

            Combinational functions in 7-Series FPGAs are synthesized as Look-Up Tables (LUTs). 
            The FPGA synthesizer maps the combinational element into LUTs and the memory element onto FFs.

            \begin{figure}[H]
                \centering
                \includegraphics[width=8.0cm]{figures/fsm.png}
                \caption{A Finite State Machine (Moore)}
                \label{fig:fsm}
            \end{figure}

            In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}.
            Shown in Figure \ref{fig:intrasite_intersite} are two possible placements for this cell net. 
            On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
            On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.

@@

            \begin{figure}[H]
                \centering
                \includegraphics[width=9.0cm]{figures/intersite_intrasite_2.png}
                \caption{
                    \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
                    \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
                    \textbf{Bottom}: The LUT-FF pair placed across different Sites.
                }
                \label{fig:intrasite_intersite}
            \end{figure}

@@

    \begin{figure*}[b]
        \centering
        \includegraphics[width=\textwidth]{figures/slicem.png}
        \caption{SLICEM Site.}
    \end{figure}

    \begin{figure*}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/slicel.png}
        \caption{SLICEL Site.}
    \end{figure}
